# Installing software and the `ebvcube` R package {#sec-installation}

::: {.callout-note appearance="simple" title="In this unit"}
By the end you will be able to:

-   confirm that your R installation meets the requirements for `ebvcube`;
-   install the `ebvcube` package and its dependencies;
-   verify your setup by loading the package and opening an example dataset.
:::

This unit prepares your machine for the rest of the tutorial. You install `ebvcube` and its dependencies, then test the installation by opening an example dataset. Once that works, you are ready to prepare your own data in [Unit 3](03-preparing-data.qmd).

## What you need first

Before installing `ebvcube`, make sure you have:

::: {.grid}

::: {.g-col-12 .g-col-md-6 .border .rounded .p-3}
**R ≥ 4.2.0** [required]{.badge .bg-secondary}

[The minimum version `ebvcube` supports. Check yours with `R.version.string`, and if it is older, update R first, or CRAN will refuse to install the package.]{.small .text-muted}
:::

::: {.g-col-12 .g-col-md-6 .border .rounded .p-3}
**RStudio** [recommended]{.badge .bg-secondary}

[A free, widely used editor for R. It is not required, but it makes it easier to run, save, and revisit the code in this tutorial. Download it [here](https://posit.co/downloads).]{.small .text-muted}
:::

:::

::: {.callout-note collapse="true" title="What about HDF5 and GDAL?"}
On Windows and macOS, you will usually not need to install **HDF5** or **GDAL** yourself. Bioconductor supplies HDF5 through `rhdf5`, while CRAN binary builds of `terra` normally include the spatial libraries they need. Linux installations can require additional system libraries; [Unit 7](07-troubleshooting.qmd) covers the common fixes.
:::

## Installing `ebvcube` from CRAN

For most users, the released version on CRAN is all you need:

```{r}
#| eval: false
install.packages("ebvcube")
```

CRAN resolves all of the *CRAN* dependencies automatically. Three dependencies, however, live on **Bioconductor** rather than CRAN, so on a fresh machine you will usually need the one-time step in the next section as well.

::: {.callout-note collapse="true" title="Other ways to install"}
Most people should use CRAN. If you need a newer feature, you can install the development version with `devtools::install_github("EBVcube/ebvcube", ref = "dev")`, or get the binary builds from the [b-cubed r-universe](https://b-cubed-eu.r-universe.dev/). These are covered in the package's [installation guide](https://github.com/EBVcube/ebvcube).
:::

## Adding the Bioconductor dependencies {#sec-bioc}

Three of `ebvcube`'s dependencies come from [Bioconductor](https://bioconductor.org/) rather than CRAN, because reading EBVCube files means reading HDF5 under the hood:

::: {.grid}

::: {.g-col-12 .g-col-md-4 .border .rounded .p-3}
`rhdf5`

[Low-level HDF5 read/write, since netCDF-4 is HDF5 underneath.]{.small .text-muted}
:::

::: {.g-col-12 .g-col-md-4 .border .rounded .p-3}
`DelayedArray`

[Lazy, on-disk arrays so large cubes do not exhaust your RAM.]{.small .text-muted}
:::

::: {.g-col-12 .g-col-md-4 .border .rounded .p-3}
`HDF5Array`

[Connects `DelayedArray` to HDF5 files.]{.small .text-muted}
:::

:::

If installing `ebvcube` reports that one of these is missing, install them with `BiocManager`:

```{r}
#| eval: false
install.packages("BiocManager")
BiocManager::install(c("rhdf5", "DelayedArray", "HDF5Array"))
```

You only need to do this once per R installation.

::: {.callout-note collapse="true" title="Which Bioconductor version?"}
`BiocManager` automatically selects the Bioconductor release that matches your version of R. Leave the version unspecified unless a package maintainer or a project requirement tells you otherwise.
:::

## Verifying your installation

First, load the package and confirm the version:

```{r}
#| eval: false
library(ebvcube)
packageVersion("ebvcube")
```

Then open the small example dataset that ships with the package. Because it is bundled, this needs no download and works offline, making it a reliable installation check:

```{r}
#| eval: false
file <- system.file("extdata", "martins_comcom_subset.nc", package = "ebvcube")

prop <- ebv_properties(file, verbose = FALSE)
prop@general$title
#> [1] "Local bird diversity (cSAR/BES-SIM)"
```

::: {.callout-important collapse="true" title="Why is the test file different from the tutorial example?"}
They are two different published datasets, which can be easy to miss.

The file bundled inside `ebvcube` is a spatial subset of **[Local bird diversity (cSAR/BES-SIM)](https://portal.geobon.org/ebv-detail?id=1)**, portal id = 1. The dataset this tutorial follows is **Global trends in biodiversity (BES-SIM cSAR-iDiv)**, portal id = 30. They share a lead author and a model family, which is why the names look so similar, but they are separate published products: id = 1 covers 1900–2015 with no scenarios, while id = 30 covers 1900, 2015 and 2050 across three SSP×RCP scenarios.

We use the bundled file here and only here, because verifying an installation should not depend on a working internet connection. The tutorial's scenario-based reference dataset is id = 30.
:::

If the title prints, your installation is working. You can then optionally download the tutorial's scenario-based reference dataset from the portal (≈ 3.3 MB). This second check requires an internet connection:

```{r}
#| eval: false
file <- ebv_download(id = 30, outputdir = tempdir(), overwrite = FALSE)
prop <- ebv_properties(file, verbose = FALSE)
prop@general$title
#> [1] "Global trends in biodiversity (BES-SIM cSAR-iDiv)"
```

::: {.callout-tip appearance="simple"}
**Checkpoint.** Your setup is complete when `library(ebvcube)` loads without error and `ebv_properties()` prints the title of the bundled dataset. Downloading id = 30 is an optional connectivity check. Once the installation check works, you are ready to prepare biodiversity data in [Unit 3](03-preparing-data.qmd).
:::
