# Validating your EBVCube before uploading {#sec-validating}

::: {.callout-note appearance="simple" title="In this unit"}
By the end you will be able to:

-   inspect the structure of your netCDF with `ebv_properties()` and `ebv_datacubepaths()`;
-   check that the cube values fall in a sensible range with `ebv_analyse()`;
-   see the data as a map with `ebv_map()` and as a temporal trend with `ebv_trend()`;
-   compare the finished file against the metadata you declared;
-   run a short pre-flight checklist before you upload.
:::

In [@sec-creating] you built an EBVCube and gave it a first structural look. This unit checks that the file is actually right before it goes to the EBV Data Portal in [@sec-uploading]. Every problem you catch here is cheaper to fix now than after a reviewer finds it.

Validation runs on four quick questions — each one function, each cheaper than a rejected upload:

::: {.grid}

::: {.g-col-12 .g-col-sm-6 .g-col-md-3 .border .rounded .p-3}
[STRUCTURE]{.small .text-success}\
Is the shape right?

[`ebv_datacubepaths()` and `ebv_properties()` list the cubes and metadata.]{.small .text-muted}
:::

::: {.g-col-12 .g-col-sm-6 .g-col-md-3 .border .rounded .p-3}
[VALUES]{.small .text-success}\
Are the numbers sane?

[`ebv_analyse()` reports min, max, mean and the valid-pixel count.]{.small .text-muted}
:::

::: {.g-col-12 .g-col-sm-6 .g-col-md-3 .border .rounded .p-3}
[VISUAL]{.small .text-success}\
Does it look right?

[`ebv_map()` and `ebv_trend()` turn the cube into a picture.]{.small .text-muted}
:::

::: {.g-col-12 .g-col-sm-6 .g-col-md-3 .border .rounded .p-3}
[METADATA]{.small .text-success}\
Does it say the truth?

[Read the attributes back and match them to what you declared.]{.small .text-muted}
:::

:::

The checks below run against a real file. So the unit is self-contained, we validate the published dataset (id = 1); when you validate your *own* cube, point `nc` at the file you created in [@sec-creating] instead:

```{r}
#| eval: false
library(ebvcube)

# Self-contained: download the published cube to a temporary directory.
# For your own file, replace this with: nc <- "path/to/your_cube.nc"
nc <- ebv_download(id = 1, outputdir = tempdir(), overwrite = FALSE)
dc <- "metric_1/ebv_cube"    # the running example has no scenario layer
```

## Check the structure

Start with the two functions you already met. `ebv_datacubepaths()` lists every cube in the file, and `ebv_properties()` reports the metadata slot by slot. Read them against what you meant to build:

```{r}
#| eval: false
ebv_datacubepaths(nc, verbose = FALSE)

props <- ebv_properties(nc, verbose = FALSE)
props@general$entity_names     # the entities you declared
props@spatial$extent           # c(xmin, xmax, ymin, ymax)
props@temporal$dates           # the time axis
```

The datacube paths should list every metric and entity you expect, the entity names should read correctly, and the extent and dates should match your data. A missing entity or an empty time axis points to a step that went wrong in creation.

## Check the values

Structure can be right while the numbers are wrong, so read a slice and look at its range. `ebv_analyse()` returns summary statistics for a cube or a subset:

```{r}
#| eval: false
stats <- ebv_analyse(nc, dc, entity = 1, timestep = 1, verbose = FALSE)
stats[c("min", "max", "mean", "n", "NAs")]
```

`n` is the number of valid pixels. If it is zero, or `NAs` covers the whole grid, the cube never received its data. A `min` or `max` far outside what your variable should hold usually signals a unit problem or a stray fill value that leaked into the data.

## See it — map and trend

Numbers only tell you so much; a picture catches problems a summary hides, like data in the wrong hemisphere or an all-empty ocean. This is where `ebvcube` shines — two one-line functions turn a cube into a figure with no plotting or projection code of your own.

::: {.panel-tabset}

#### Map a slice

`ebv_map()` draws a single time step as a classed choropleth, optionally overlaying country borders. One look confirms the data sit where they should:

```{r}
#| eval: false
#| fig-cap: "Relative change in bird species, first time step (global)."
ebv_map(nc, dc, entity = 1, timestep = 1, classes = 9, verbose = FALSE)
```

#### Plot the trend

`ebv_trend()` collapses the spatial dimensions to one value per time step and draws the trajectory. It does double duty — it plots **and** returns the underlying numbers, so a flat line, a break, or an implausible jump is easy to spot and the values are yours to keep:

```{r}
#| eval: false
#| fig-cap: "Mean relative change in bird species over time."
trend_vals <- ebv_trend(nc, dc, entity = 1, method = "mean", verbose = FALSE)
trend_vals
```

:::

::: {.callout-note appearance="simple" title="If the map looks shifted by 180°"}
The file is using the `[0, 360]` longitude convention while the basemap expects `[-180, 180]`. Read the slice and wrap it before plotting:

```r
r <- ebv_read(nc, dc, entity = 1, timestep = 1, type = "r")
r <- terra::rotate(r)
```
:::

## Compare against your metadata

Finally, confirm the file says what you told it to say. Read the attributes back and check them against the metadata form (or the `ebv_metadata()` call) you used to create it:

```{r}
#| eval: false
props@general$title
ebv_properties(nc, dc, verbose = FALSE)@ebv_cube$units
```

The title, units, entity count, and time coverage should all match your metadata. If a unit or description is wrong, fix it in place with `ebv_attribute()` from [@sec-creating] rather than rebuilding the file.

## A pre-flight checklist

Before you move to the portal, walk through this list. If every box ticks, the file is ready to upload:

- [ ] The datacube paths list every metric and entity you expect.
- [ ] Entity names, dates, extent, and resolution match your metadata.
- [ ] Units and the cube description read correctly.
- [ ] A map of at least one slice looks right, with no 180° shift.
- [ ] Value ranges are plausible and the valid-pixel count (`n`) is not zero.

The portal runs its own validation when you submit, but clearing these first saves a round trip. Panoply, introduced in [@sec-creating], is a quick way to eyeball the structure outside R if you want a second opinion.

::: {.callout-tip appearance="simple"}
**Checkpoint.** Your file's structure, values, map, and trend all match the dataset you meant to create, and every box in the pre-flight checklist is ticked. If so, you are ready to upload it to the EBV Data Portal in [@sec-uploading].
:::
